---
title: "Capstone data science project"
subtitle: "Geochemical populations with the WA Department of Mines, Industry, Safety and Regulation"
author: "Helena Morton (22598448)"
output:
  html_notebook: 
    toc: true
    theme: united
  pdf_document: 
    toc: false
    number_sections: no
  html_document: 
    toc: true
  word_document: 
    toc: true
date: '`r format(Sys.time(), "%d %B %Y, %H:%M ")`'
---

> The following notebook is an exploratory data analysis (EDA) and hierarchial clustering of the data provided for our Capstone data science project. The project: 

# Notebook preamble

```{r Document preamble, warning=FALSE, results='hide',message=FALSE,include=FALSE}

# List of packages required
list_of_packages <- c("ggplot2", "tidyverse","ggpubr","dplyr","lubridate","Matrix","factoextra","graphics","mvdalab","pls","itertools","corrplot")

# install packages for new users
new_packages <- list_of_packages[!(list_of_packages %in% installed.packages()[,"Package"])]
if(length(new_packages)){
install.packages(new_packages)
}

# import/load libraries
suppressPackageStartupMessages(lapply(c(list_of_packages),library, character.only = TRUE))
```

> The following notebook is an exploratory data analysis (EDA) and hierarchial clustering of the data provided for our Capstone data science project. The project: 

> DATA NOTES:

> * There are $n$ = 1351 (was expecting 600) and $p$ (variable) = 95. 
> * Of which $n_{basalt}$ = 1051 are and $n_{dolerite}$ = 300
> * Will work will both `Rocktypes`

> * Need to decide which variables/columns ($p$) are needed/essential since dataset is large
> * Check for missing values (NA's) and decide what to do with these samples
> * Apparently the NA's could also be so low that they are undetected by the analysis

> * $n$ = 6 variables that are 'chr'data type, $n$ = 89 that are 'num' data type
> * consider data normalisation for num dt?
> * What to do about outliers?

# Import & load dataset

```{r import data}
# import/load data
geo_data <- read.csv("kalgoolie-geochemistry.csv") # name of dataset
# list the columns in dataset
colnames(geo_data)
cat('\nThere are',nrow(geo_data),'rows/ oberservations (n)','and',ncol(geo_data),'columns/ variables (p) in this dataset\n\n')
```

# Exploratory data analysis (EDA) 

> Exploration of data, cleaning and transformations

```{r }
geo_columns <- cbind(seq(1,ncol(geo_data),1),names(geo_data)) %>% `colnames<-`(c("Number","Names"))
as.table(geo_columns)

geo_columns[,2]
```


Number Names                  
  1      ï..SampleID            
  2      Rocktype               
  3      Source                 
  4      Publication1           
  5      Publication2           
  6      Project                
  7      Dlat                   
  8      Dlong                  
  9      Sample_type            
  10     Core                   
  11     Depth_m_from           
L 12     Depth_m_to             
M  13     Description_Other_notes
N  14     X                      
O  15     SiO2_pct               
P  16     TiO2_pct               
Q  17     Al2O3_pct              
R  18     Fe2O3T_pct             
S  19     MgO_pct                
T  20     MnO_pct                
U  21     CaO_pct                
V  22     Na2O_pct               
W  23     K2O_pct                
X  24     P2O5_pct               
Y  25     SO3_pct                
Z  26     LOI_pct                
A1 27     Cl_pct                 
B1 28     Total_calc             
C1 29     Total_LOI              
D1 30     Total_LOI_SO3          
E1 31     aSiO2_pct              
F1 32     aTiO2_pct              
G1 33     aAl2O3_pct             
H1 34     aFe2O3T_pct            
I1 35     aMgO_pct               
J1 36     aMnO_pct               
K1 37     aCaO_pct               
L1 38     aNa2O_pct              
M1 39     aK2O_pct               
N1 40     aP2O5_pct              
O1 41     aTotal                 
P1 42     FeOT_pct               
Q1 43     Mg                     
R1 44     Na20_K2O               
S1 45     K2O_Na2O               
T1 46     ASI                    
U1 47     Eu_Eu                  
V1 48     Au_ppb                 
W1 49     Pt_ppb                 
X1 50     Pd_ppb                 
Y1 51     Re_ppm                 
Z1 52     Ag_ppm                 
A2 53     F_ppm                  
B2 54     Bi_ppm                 
C2 55     Sn_ppm                 
D2 56     Mo_ppm                 
E2 57     Be_ppm                 
F2 58     Cs_ppm                 
G2 59     Ba_ppm                 
H2 60     Rb_ppm                 
I2 61     Sr_ppm                 
J2 62     Pb_ppm                 
K2 63     Th_ppm                 
L2 64     U_ppm                  
M2 65     Nb_ppm                 
N2 66     Ta_ppm                 
O2 67     Zr_ppm                 
P2 68     Hf_ppm                 
Q2 69     Y_ppm                  
R2 70     Ga_ppm                 
S2 71     La_ppm                 
T2 72     Ce_ppm                 
U2 73     Pr_ppm                 
V2 74     Nd_ppm                 
W2 75     Sm_ppm                 
X2 76     Eu_ppm                 
Y2 77     Gd_ppm                 
Z2 78     Tb_ppm                 
A3 79     Dy_ppm                 
B3 80     Ho_ppm                 
C3 81     Er_ppm                 
D3 82     Tm_ppm                 
E3 83     Yb_ppm                 
F3 84     Lu_ppm                 
G3 85     Sc_ppm                 
H3 86     Ni_ppm                 
I3 87     V_ppm                  
J3 88     Cd_ppm                 
K3 89     Cr_ppm                 
L3 90     Cu_ppm                 
M3 91     Zn_ppm                 
N3 92     Sb_ppm                 
O3 93     As_ppm                 
P3 94     Ge_ppm                 
Q3 95     Ti_ppm      

```{r reduce DF}
# chosen columns to remove
remove_cols <- c(1:26,28:30,41,46,47)
remove_cols1 <- c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,
          17,18,19,20,21,22,23,24,25,26,28,29,30,41,46,47)

# could use this to remove columns based on number
#new_geo <- geo[-c(remove_cols)]

# Remove columns from dataset
reduce_geo <- geo_columns[,1] %in% remove_cols

reduce <- cbind(reduce_geo,geo_columns)  %>% 
  as_tibble(c.name_repair='unique') %>% 
  filter(reduce_geo==FALSE)

# New dataframe
geo_chemical_data <- geo_data[colnames(geo_data) %in% reduce$Names]

# all the column's removed
cat("You want to remove the following columns:\n\n")
removed <- cbind(reduce_geo, geo_columns)  %>% 
  as_tibble(c.name_repair='unique') %>% 
  filter(reduce_geo==TRUE)
as.matrix(removed[-1])

# selected columns
cat("\nYou want to keep the following columns:\n\n")
as.matrix(names(geo_chemical_data))
# New Dataset's dimensions
cat('\n\nThere are',nrow(geo_chemical_data),'rows/ oberservations (n)','and',ncol(geo_chemical_data),'columns/ variables (p) in this dataset\n\n')
```

Since there are many NA's/ missing values - although it is noted that these could be measurements that are too low to be detected by the analytical machinery - they are replaced with zeros's here instead of random numbers from the distribution.

```{r check for missing values}
# The n of Na's in df
message("Missing values detection in new dataframe (DF)\n")
sapply(geo_chemical_data,function(x) sum((is.na(x))))
```

```{r data transformations}
# fill NA's with zero
geochemistry_zeros <- geo_chemical_data
geochemistry_zeros[is.na(geochemistry_zeros)] <- 0
# remove NA's
geochemistry_omit_na <- na.omit(geo_chemical_data)
# log transformation of data
geochemistry_omit_na <- geo_chemical_data %>% na.omit %>% log10
# I for compositional data like geochemistry
ilr(geo_chemical_data)

ilr( x , V = ilrBase(x) ,...)
          ilrInv( z , V = ilrBase(z=z),..., orig=gsi.orig(z))

```



```{r}
sapply(geo_chemical,function(x) sum((is.na(x))))
```


```{r}
# Plots of variables
for (variable in geo_chemical_data){
  element_names <- names(geo_chemical_data)[variable]
  par(mfrow=c(2,2))
  plot(variable,col="seagreen",ylab = element_names[+1], main = element_names[+1], xlab = "frequency")
  boxplot(variable, horizontal = F, col = "steel blue",border = "brown",notch = TRUE, main=element_names[+1], ylab= "measurement")
  hist(variable,col="seagreen", main = element_names[+1], xlab= "measurement")
}
```


```{r scatter plots of the variables}

ggplot() + geom_point(aes(x=aTiO2_pct,y=Th_ppm), data = geo_chemical_data) + geom_smooth()

ggplot() + geom_point(aes(x=aTiO2_pct,y=La_ppm), data = geo_chemical_data) + geom_smooth()

ggplot() + geom_point(aes(x=aTiO2_pct,y=Zr_ppm)) + geom_smooth()

ggplot() + geom_point(aes(x=Th_ppm,y=Nb_ppm)) + geom_smooth()

ggplot() + geom_point(aes(x=Zr_ppm,y=La_ppm), data = geo_chemical_data) + geom_smooth()

ggplot() + geom_point(aes(x=Th_ppm,y=Zr_ppm), data = geo_chemical_data) + geom_smooth()

ggplot() + geom_point(aes(x=Mg,y=La_ppm), data = geo_chemical_data) + geom_smooth()

ggplot() + geom_point(aes(x=Mg,y=aAl2O3_pct), ddata = geo_chemical_data) + geom_smooth()

ggplot() + geom_point() + geom_smooth()

#ggplot(new_DF, aes(x=Mg,y=aTiO3_pct)) + geom_point() + geom_smooth()

ggplot() + geom_point(aes(x=Dlat,y=Dlong)) + geom_smooth()
```


```{r Scatterplot function}

scatter_plot <- function(mydf, myxcol, myycol, mytitle) {
   ggplot2::ggplot(data = mydf, aes_string(x=reorder({{ myxcol }}, 
      {{ myycol }}), y= {{ myycol }})) + 
    geom_point(color = "black", fill="#0072B2") +
    xlab(colnames(myxcol)) +
    ylab(colnames(myxcol)) +
    ggtitle(mytitle) 
}

scatter_plot(geo_chemical_data, aTiO2_pct, Th_ppm, "scatterplot")
```

# Correlations 

```{r Correlation matrix}
# correlations between variables
round(cor(geo_chemical_data),3) # large matrix
```

```{r fig.height = 12, fig.width = 12}

# save correlation matrix as a variable
chem_cor_matrix <- round(cor(geo_chemical_data),3)
# order the correlations from strongest to weakest
chem_cor_matrix_all <- chem_cor_matrix %>%
  as.data.frame() %>%
  mutate(var1 = rownames(.)) %>%
  gather(var2, value, -var1) %>%
  arrange(desc(value)) %>%
  group_by(value) %>%
  filter(row_number() == 1.000) 

# the 'strong' > 0.85 correlations in the DF are they related? there are no correlations < -0.85
strong_chem_cor_neg <- chem_cor_matrix_all %>%  filter(value < -0.80) # high correlation limit declared
strong_chem_cor_pos <- chem_cor_matrix_all %>%  filter(value > 0.85) # high correlation limit declared

chem_cor_df <- strong_chem_cor_pos %>% as.matrix(rownames.force = TRUE) %>% as_data_frame
new_strong_cor <- chem_cor_df %>% dplyr::rename(Correlation_r = "value")
new_strong_cor
# from n=64 now there are n=32
new_chem <- new_DF[colnames(new_DF) %in% c(chem_cor_df$var1,chem_cor_df$var2)]
#names(new_chem)
```

NOTE: build plots based on big pair plots


```{r Strong correlation plots}

```


```{r Correlation plot, fig.height = 15, fig.width = 12}
ggcorr(new_DF, label=T,digits=2,low="pink",mid="steelblue",high="seagreen",label_alpha = TRUE, nbreaks = 10)
corrplot(cor(new_DF))
```

NOTES:

Bold blue = strongly correlated (positive) ~0.8

Bold red = strongly correlated (negative) ~ -0.8

select strong correlations between main elements and strong correlations between main/trace elements.

# Hierarchial Clustering

```{r cluster functions}
# Hierarchial functions

set.seed(22598448) # for reproducibility
# FUNCTION Hierarchial clustering with three distance measures
Hierachial_cluster <- function(DF,distance_method,linkage_method,K,plot_title){
  
  # scale DF, already scaled? select condition 
  DF <- DF %>% dplyr::select_if(is.numeric) %>% as.data.frame #%>% scale
  # distance matrix # hclust & linkage method
  hclust_method<-hclust(dist(DF,method =distance_method),method=linkage_method)
  # dendogram
  dendogram <- as.dendrogram(hclust_method)
  # plot the dendrogram with color-coded groups
  col_dendogram <- color_branches(dendogram, h = max(hclust_method$height),k=K,lwd=2)
  plot(col_dendogram, lwd = 2, main = plot_title)
  # plot clusters
  abline(h = max(hclust_method$height), col = 'green', lwd = 2)
  hclust_method
}
```

### Transposing of the dataframe 

NOTE: won't work with NA's

```{r}
transposed_DF <- data.frame(t(geo_chemical_data))

# Hierachial clustering with k=1,...9 for MANHATTAN distance & COMPLETE linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(transposed_DF),"manhattan","complete",K=i,"Cluster dendogram: complete and manhattan distance")
  hc_mc
}

# Hierachial clustering with k=1,...9 for MANHATTAN distance & SIN== linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(transposed_DF),"manhattan","single",K=i,"Cluster dendogram: single and manhattan distance")
  hc_mc
}

# Hierachial clustering with k=1,...9 for MANHATTAN distance & COMPLETE linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(transposed_DF),"manhattan","average",K=i,"Cluster dendogram: average and manhattan distance")
  hc_mc
}
```



```{r}
transposed_DF <- data.frame(t(geo_chemical_data))
t_chemical_scaled <- as_tibble(scale(transposed_DF))

# Hierachial clustering with k=1,...9 for MANHATTAN distance & COMPLETE linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(t_chemical_scaled),"euclidean","complete",K=i,"Cluster dendogram: complete and manhattan distance")
  hc_mc
}

# Hierachial clustering with k=1,...9 for MANHATTAN distance & SINGLE= linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(t_chemical_scaled),"euclidean","single",K=i,"Cluster dendogram: single and manhattan distance")
  hc_mc
}

# Hierachial clustering with k=1,...9 for MANHATTAN distance & average linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(t_chemical_scaled),"euclidean","average",K=i,"Cluster dendogram: average and manhattan distance")
  hc_mc
}
```


### Scaled, no NA's

```{r}
chemical_scaled <- as_tibble(scale(geo_chemical_data))

# Hierachial clustering for euclidean distance & COMPLETE linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(chemical_scaled),"euclidean","complete",K=i,"Cluster dendogram: complete and manhattan distance")
  hc_mc
}

# Hierachial clustering with for euclidean distance & SINGLE= linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(chemical_scaled),"euclidean","single",K=i,"Cluster dendogram: single and manhattan distance")
  hc_mc
}

# Hierachial clustering with  for euclidean distance & average linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(chemical_scaled),"euclidean","average",K=i,"Cluster dendogram: average and manhattan distance")
  hc_mc
}

# Hierachial clustering with  for MANHATTAN distance & COMPLETE linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(transposed_DF),"manhattan","complete",K=i,"Cluster dendogram: complete and manhattan distance")
  hc_mc
}

# Hierachial clustering with  for MANHATTAN distance & single linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(transposed_DF),"manhattan","single",K=i,"Cluster dendogram: single and manhattan distance")
  hc_mc
}

# Hierachial clustering with for MANHATTAN distance & average linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(transposed_DF),"manhattan","average",K=i,"Cluster dendogram: average and manhattan distance")
  hc_mc
}
```

### Unscaled data, no NA's

```{r}
chemical_scaled <- as_tibble(scale(geo_chemical_data))

# Hierachial clustering for euclidean distance & COMPLETE linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(geo_chemical_data),"euclidean","complete",K=i,"Cluster dendogram: complete and manhattan distance")
  hc_mc
}

# Hierachial clustering with for euclidean distance & SINGLE= linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(geo_chemical_data),"euclidean","single",K=i,"Cluster dendogram: single and manhattan distance")
  hc_mc
}

# Hierachial clustering with  for euclidean distance & average linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(geo_chemical_data),"euclidean","average",K=i,"Cluster dendogram: average and manhattan distance")
  hc_mc
}

# Hierachial clustering with  for MANHATTAN distance & COMPLETE linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(geo_chemical_data),"manhattan","complete",K=i,"Cluster dendogram: complete and manhattan distance")
  hc_mc
}

# Hierachial clustering with  for MANHATTAN distance & single linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(geo_chemical_data),"manhattan","single",K=i,"Cluster dendogram: single and manhattan distance")
  hc_mc
}

# Hierachial clustering with for MANHATTAN distance & average linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(geo_chemical_data),"manhattan","average",K=i,"Cluster dendogram: average and manhattan distance")
  hc_mc
}
```

### unscaled data

```{r}
chemical_scaled <- as_tibble(scale(geo_chemical_data))

# Hierachial clustering for euclidean distance & COMPLETE linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(geo_chemical_data),"euclidean","complete",K=i,"Cluster dendogram: complete and manhattan distance")
  hc_mc
}

# Hierachial clustering with for euclidean distance & SINGLE= linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(geo_chemical_data),"euclidean","single",K=i,"Cluster dendogram: single and manhattan distance")
  hc_mc
}

# Hierachial clustering with  for euclidean distance & average linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(geo_chemical_data),"euclidean","average",K=i,"Cluster dendogram: average and manhattan distance")
  hc_mc
}

# Hierachial clustering with  for MANHATTAN distance & COMPLETE linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(geo_chemical_data),"manhattan","complete",K=i,"Cluster dendogram: complete and manhattan distance")
  hc_mc
}

# Hierachial clustering with  for MANHATTAN distance & single linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(geo_chemical_data),"manhattan","single",K=i,"Cluster dendogram: single and manhattan distance")
  hc_mc
}

# Hierachial clustering with for MANHATTAN distance & average linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(na.omit(geo_chemical_data),"manhattan","average",K=i,"Cluster dendogram: average and manhattan distance")
  hc_mc
}
```

### unscaled data, imputed zero's inplace of NA's

```{r}

# Hierachial clustering for euclidean distance & COMPLETE linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(geochemistry_zeros,"euclidean","complete",K=i,"Cluster dendogram: complete and manhattan distance")
  hc_mc
}

# Hierachial clustering with for euclidean distance & SINGLE= linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(geochemistry_zeros,"euclidean","single",K=i,"Cluster dendogram: single and manhattan distance")
  hc_mc
}

# Hierachial clustering with  for euclidean distance & average linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(geochemistry_zeros,"euclidean","average",K=i,"Cluster dendogram: average and manhattan distance")
  hc_mc
}

# Hierachial clustering with  for MANHATTAN distance & COMPLETE linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(geochemistry_zeros,"manhattan","complete",K=i,"Cluster dendogram: complete and manhattan distance")
  hc_mc
}

# Hierachial clustering with  for MANHATTAN distance & single linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(geochemistry_zeros,"manhattan","single",K=i,"Cluster dendogram: single and manhattan distance")
  hc_mc
}

# Hierachial clustering with for MANHATTAN distance & average linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(geochemistry_zeros,"manhattan","average",K=i,"Cluster dendogram: average and manhattan distance")
  hc_mc
}
```

### scaled data, imputed zero's inplace of NA's

```{r}

# Hierachial clustering for euclidean distance & COMPLETE linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(as_tibble(scale(geochemistry_zeros)),"euclidean","complete",K=i,"Cluster dendogram: complete and manhattan distance")
  hc_mc
}

# Hierachial clustering with for euclidean distance & SINGLE= linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(as_tibble(scale(geochemistry_zeros)),"euclidean","single",K=i,"Cluster dendogram: single and manhattan distance")
  hc_mc
}

# Hierachial clustering with  for euclidean distance & average linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(as_tibble(scale(geochemistry_zeros)),"euclidean","average",K=i,"Cluster dendogram: average and manhattan distance")
  hc_mc
}

# Hierachial clustering with  for MANHATTAN distance & COMPLETE linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(as_tibble(scale(geochemistry_zeros)),"manhattan","complete",K=i,"Cluster dendogram: complete and manhattan distance")
  hc_mc
}

# Hierachial clustering with  for MANHATTAN distance & single linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(as_tibble(scale(geochemistry_zeros)),"manhattan","single",K=i,"Cluster dendogram: single and manhattan distance")
  hc_mc
}

# Hierachial clustering with for MANHATTAN distance & average linkage
for (i in 2:3){
  hc_mc <-Hierachial_cluster(as_tibble(scale(geochemistry_zeros))\,"manhattan","average",K=i,"Cluster dendogram: average and manhattan distance")
  hc_mc
}
```








